---
title: "HeLa Adaptability"
author: "Richard R. Evans"
date: "May 15, 2017"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Study Description

From Adam Lauring:

We examined the dynamics of adaptive evolution by passage of both wild type (WT) and 3D G64S virus populations on HeLa or PVR-3T3 cells. 

Starting from the same stock of either WT or 3D G64S, we passed 5 replicate lineages of each virus for 20 passages (HeLa) or 12 passages (PVR-3T3). 

We measured the fitness of the passage 0 stocks and every 4th and 5th passage of the evolved lineages for the HeLa experiment and PVR-3T3 experiments, respectively, by competition against a tagged WT reference in triplicate.

For the HeLa experiment, then there are 5 replicate lineages, 4 time points per lineage (in addition to the starting passage 0 stock), and 3 replicate fitness measurements per lineage-time point. 

For the PVR-3T3 experiment, then there are 5 replicate lineages, 3 time points per lineage (in addition to the starting passage 0 stock), and 3 replicate fitness measurements per lineage-time point.

# Libraries

```{r, library loading}
library(tidyverse)
library(magrittr)
library(lme4)
library(lmerTest)
library(pander)
library(reshape2)
library(sjPlot)
```

# Load Data

```{r}
# setwd("~/Documents/R/hela_adaptability/")
hela <- read.table(file = "HeLa_Adaptability_20.txt", header = TRUE)
cell_3t3 <- read.table(file = "3T3_Adaptability_P12_05302017.txt", header = TRUE)
```

# Data Processing

```{r}
hela_clean <- hela %>%
  select(-fitness) %>%
  mutate(virus = factor(virus, levels = c(0, 1),
                        labels = c("Virus 0"," Virus 1")),
         repID = factor(repID, levels = 1:5,
                        labels = paste(rep("repID:", 5), 1:5, sep = "")),
         measID = factor(measID)) %>%
  arrange(virus, time)

cell3t3_clean <- cell_3t3 %>%
  select(-fitness) %>%
  mutate(virus = factor(virus, levels = c(0, 1),
                        labels = c("Virus 0"," Virus 1")),
         repID = factor(repID, levels = 1:5,
                        labels = paste(rep("repID:", 5), 1:5, sep = "")),
         measID = factor(measID)) %>%
  arrange(virus, time)
```

## HeLa data preview

```{r}
head(tbl_df(hela_clean))
```
.

.

.
```{r}
tail(tbl_df(hela_clean))
```

## 3T3 Cell Preview

```{r}
head(tbl_df(cell3t3_clean))
```

# Data Exploration

## HeLa Cells

```{r}
pander(sapply(hela_clean[,c(1:4)], table))
```

There are 119 $virus_{0}$ and 117 $virus_{1}$. The missing two are from missing values during lab testing, they have been removed from the analysis.

Likewise, there are 3 time points missing from t=10 and 1 time point missing from t=20. Two time points are missing from $repID_{1}$ and $repID_{2}$. Two time points are missing from $measID_{1}$ and $measID_{3}$

By replicate measure and virus type, the effect of measurement at time point 0 is always the same for a particular type of virus. I don't think the intercept matters at all in assessing virus 1 vs virus 0. This can be seen in the following plot.

```{r}
hela_clean %>%
  filter(as.numeric(measID) <= 3) %>%
  ggplot(aes(x = time, y = logfitness, color = measID)) %>%
  add(geom_point(stat = "identity", position = "identity",
                 alpha = 0.7, size = 3)) %>%
  add(geom_smooth(method = "lm", formula = y ~ x, se = FALSE)) %>%
  add(facet_wrap(~virus + repID, nrow = 2)) %>%
  add(labs(x = "time",
           y = "log(Fitness)")) %>%
  add(theme_bw())

hela_clean %>%
  ggplot(aes(x = time, y = logfitness, color = measID)) %>%
  add(geom_point(stat = "identity", position = "identity",
                 alpha = 0.7, size = 3)) %>%
  add(geom_smooth(method = "lm", formula = y ~ x, se = FALSE)) %>%
  add(facet_wrap(~virus + repID, nrow = 2)) %>%
  add(labs(x = "time",
           y = "log(Fitness)")) %>%
  add(theme_bw())
```

Without t = 0

```{r}
hela_clean %>%
  filter(as.numeric(measID) <= 3 & time != 0) %>%
  ggplot(aes(x = time, y = logfitness, color = measID)) %>%
  add(geom_point(stat = "identity", position = "identity",
                 alpha = 0.7, size = 3)) %>%
  add(geom_smooth(method = "lm", formula = y ~ x, se = FALSE)) %>%
  add(facet_wrap(~virus + repID, nrow = 2)) %>%
  add(labs(x = "time",
           y = "log(Fitness)")) %>%
  add(theme_bw())
```

## 3T3 Cells

```{r}
pander(sapply(cell3t3_clean[,c(1:4)], table))
```

```{r}
cell3t3_clean %>%
  filter(as.numeric(measID) <= 3) %>%
  ggplot(aes(x = time, y = logfitness, color = measID)) %>%
  add(geom_point(stat = "identity", position = "identity",
                 alpha = 0.7, size = 3)) %>%
  add(geom_smooth(method = "lm", formula = y ~ x, se = FALSE)) %>%
  add(facet_wrap(~virus + repID, nrow = 2)) %>%
  add(labs(x = "time",
           y = "log(Fitness)")) %>%
  add(theme_bw())
```

# Models

## Naive Model (HeLa Cell)
```{r}
naive.mod <- glm(logfitness ~ virus + time + virus * time,
                 data = hela_clean)

summary(naive.mod)
```

This is a very silly model. There is no nesting involved and clearly a lot of duplicated values. This would severely reduce our standard errors and that is why you see very small p-values.

```{r}
naive.mod <- glm(logfitness ~ virus + time + virus * time,
                 data = distinct(hela_clean, virus, time, logfitness))

summary(naive.mod)
```

This version, with distinct values of logfitness, time, and virus is not really a good idea either but it would reflect the actual experiment while ignoring the hierarchical structure of the experiment.

## Hierarchical Models

### HeLa Cell

#### Frequentist

This is a three-level model with time nested within measID nested within repID.

```{r}
mod.1 <- lmer(logfitness ~ virus * time + (time|measID/repID),
              data = hela_clean,
              REML = FALSE)

summary(mod.1)
# plot(mod.1)
```

For comparison we will run a model without the measID/repID to include only an interaction between measID:repID and a main effect for measID. The former reduces to `virus + time + virus * time + (time|measID) + (time|repID) + (time|measID) * (time|repID)`.

```{r}
mod.2 <- lmer(logfitness ~ virus * time + (time|repID:measID),
              data = hela_clean,
              REML = FALSE)

summary(mod.2)

anova(mod.1, mod.2)
```

There is essentially no difference between the two. To simplify you could take mod.2.

Plotting frequentist mixed models can be dubious. You can use the visreg package but it shouldn't be trusted for inference since you can't easily estimate a standard error from predictions (fuzzy p-values again.)

```{r}
visreg::visreg(mod.1, "time", by = "virus", overlay = TRUE)
visreg::visreg(mod.1, "repID", by = "virus")
visreg::visreg(mod.1, "measID", by = "virus")
```

Measure ID looks very strange because of the 12 measurements at t = 0 causing imbalance.

You can see this more easily in the new sjPlot package:

```{r}
sjp.lmer(mod.1)
```

These are the BLUPs, the "Best Linear Unbiased Predictors" for every level of the random effects. at any of the $measID = \{1, 2, 3\}$ you can see the predictons vary, however at every value of $measID = \{4, ..., 12\}$ they are almost identical.

```{r}
sjp.lmer(mod.1, type = "fe", sort = TRUE)
```

The virus by time interaction is really only marginally significant but that is mostly due to the differing intercept.

```{r}
sjp.lmer(mod.1, type = "rs.ri", 
         vars = "virus", sample.n = 12, show.legend = TRUE)
```
These are the random slopes for time by measure ID, where they basically don't vary for measID 4:12 but do vary for 1:3, sort of.
```{r}
sjp.lmer(mod.1, type = "rs.ri", 
         vars = "virus", sample.n = c(1,2,3), show.legend = TRUE)
```

#### Bayesian

I ran the following MCMC model and then saved it for markdown rendering purposes. I use BRMS here for ease of use, if you want to try it yourself it may take some time to set-up. First you must install Rtools, then STAN and then brms. It's a process.

```{r}
library(brms)
# brm.1 <- brm(logfitness ~ virus * time + (time|measID/repID),
#              data = hela_clean)
# 
# saveRDS(brm.1, file = "~/Documents/R/hela_adaptability/brm.rds")
brm.1 <- readRDS(file = "brm.rds")

summary(brm.1)
plot(brm.1)
marginal_effects(brm.1)
```

Bayesian modelling of the same experiment arrives at a very similar conclusion, namely there is a difference between virus 1 and virus 2 over time but only in starting point, they exhibit the same growth patterns (slopes).

### 3T3 Cell

#### Frequentist

This is a three-level model with time nested within measID nested within repID.

```{r}
mod.1 <- lmer(logfitness ~ virus * time + (time|measID/repID),
              data = cell3t3_clean,
              REML = FALSE)

summary(mod.1)
# plot(mod.1)
```

```{r}
visreg::visreg(mod.1, "time", by = "virus", overlay = TRUE)
visreg::visreg(mod.1, "repID", by = "virus")
visreg::visreg(mod.1, "measID", by = "virus")
```

#### Bayesian

```{r}
# brm.3t3 <- brm(logfitness ~ virus * time + (time|measID/repID),
#                data = cell3t3_clean)
# 
# saveRDS(brm.3t3, file = "~/Documents/R/hela_adaptability/brm_3t3.rds")
brm.3t3 <- readRDS(file = "brm_3t3.rds")

summary(brm.3t3)
plot(brm.3t3)
marginal_effects(brm.3t3)
```

# Session Info

```{r}
sessionInfo()
```

